---
title: "Classification Techniques: Linear and Quadratic Discriminant Analysis, Logistic Regression"
author: "Liam Abbott"
output: html_document
---
&nbsp;  

###Summary  

The olive dataset analyzed in this report contains data on the chemical composition of 575
Italian olive oils, measured by the percentage composition of eight different fatty acids found in
the lipid fraction of the oil. Each olive oil has an associated Region and Area label indicating
the respective region and area of origin for that particular olive oil.   

In this analysis, we conduct linear and quadratic discriminant analyses and multivariate logistic 
regression with the goal of assigning the correct region and area labels to each olive oil based on the
composition of the fatty acids found in the oil.   

The LDA, QDA, and logistic regression models are trained on a dataset consisting of the first 90% of 
observations from each of the nine areas represented in the dataset. The models are then tested on a 
test dataset consisting of the remaining 10% of observations from each of the nine areas. The training 
dataset contains 505 observations and the test dataset contains 70 observations.  

```{r, echo=FALSE, warning=FALSE, message=FALSE}
list.of.packages = c("ggplot2",
                     "gridExtra",
                     "xtable",
                     "nnet",
                     "MASS",
                     "scales",
                     "GGally")
new.packages = list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) {
  install.packages(new.packages)
}
for (i in 1:length(list.of.packages)) {
  require(list.of.packages[i], character.only=TRUE)
}
```

While the LDA, QDA, and multivariate logistic regression models all performed well when classifying olive 
oils by region of origin, LDA and QDA clearly outperformed multivariate logistic regression when classifying olive oils by area of origin. The training and test error rates for each classification model and task are given below:    

```{r, echo=FALSE, results='asis'}
#read in raw data
olive = read.table('olive.txt', header=T, sep="")

index = sort.int(olive$Area, index.return=T)$ix
olive = olive[index,]

#make region and area variables factors
olive$Region = as.factor(olive$Region)
olive$Area = as.factor(olive$Area)

#separate into training and test sets
n_obs = dim(olive)[1]
n_cols = dim(olive)[2]
M = array(0, c(9, 1))
for (i in seq(1, n_obs))
{
  t = olive$Area[i]
  M[t] = M[t]+1
}
L = floor(M*0.9)
M_cumul = c(0, cumsum(M))
select = NULL
for (i in seq(1,9))
{
  a = M_cumul[i]+1
  b = a+L[i]-1
  select = c(select, seq(a, b))
}
train = olive[select,]
test = olive[-select,]
n_train = dim(train)[1]
n_test = dim(test)[1]
```


```{r, echo=FALSE, warning=FALSE, results='hide'}
#Region linear discriminant analysis
lda_region_fit = lda(Region~Oleic+Eicosenoic, data=train)
region_train_lda_preds = predict(lda_region_fit)$class
region_train_actual = train[,1]
region_test_lda_preds = predict(lda_region_fit, newdata=test)$class
region_test_actual = test[,1]

#Region quadratic discriminant analysis
qda_region_fit = qda(Region~Oleic+Eicosenoic, data=train)
region_train_qda_preds = predict(qda_region_fit)$class
region_test_qda_preds = predict(qda_region_fit, newdata=test)$class

#Region multivariate logisitc regression analysis
log_region_fit = multinom(Region~Oleic+Eicosenoic, data=train)
region_train_log_preds = predict(log_region_fit, newdata=train, "class")
region_test_log_preds = predict(log_region_fit, newdata=test, "class")
```


```{r, echo=FALSE, warning=FALSE, results='hide'}
#Area linear discriminant analysis
lda_area_fit = lda(Area~.-Region, data=train)
area_train_lda_preds = predict(lda_area_fit)$class
area_train_actual = train[,2]
area_test_lda_preds = predict(lda_area_fit, newdata=test)$class
area_test_actual = test[,2]

#Area quadratic discriminant analysis
qda_area_fit = qda(Area~.-Region, data=train)
area_train_qda_preds = predict(qda_area_fit)$class
area_test_qda_preds = predict(qda_area_fit, newdata=test)$class

#Area multivariate logistic regression
log_area_fit = multinom(Area~.-Region, data=train)
area_train_log_preds = predict(log_area_fit, newdata=train, "class")
area_test_log_preds = predict(log_area_fit, newdata=test, "class")
```


```{r, echo=FALSE, results='asis'}
region_lda_train_error = length(region_train_actual[region_train_actual != region_train_lda_preds])/n_train
region_lda_test_error = length(region_test_actual[region_test_actual != region_test_lda_preds])/n_test

region_qda_train_error = length(region_train_actual[region_train_actual != region_train_qda_preds])/n_train
region_qda_test_error = length(region_test_actual[region_test_actual != region_test_qda_preds])/n_test

region_log_train_error = length(region_train_actual[region_train_actual != region_train_log_preds])/n_train
region_log_test_error = length(region_test_actual[region_test_actual != region_test_log_preds])/n_test

area_lda_train_error = length(area_train_actual[area_train_actual != area_train_lda_preds])/n_train
area_lda_test_error = length(area_test_actual[area_test_actual != area_test_lda_preds])/n_test

area_qda_train_error = length(area_train_actual[area_train_actual != area_train_qda_preds])/n_train
area_qda_test_error = length(area_test_actual[area_test_actual != area_test_qda_preds])/n_test

area_log_train_error = length(area_train_actual[area_train_actual != area_train_log_preds])/n_train
area_log_test_error = length(area_test_actual[area_test_actual != area_test_log_preds])/n_test

df_temp = data.frame("Region Training"=percent(round(c(region_lda_train_error, region_qda_train_error,region_log_train_error), 4)),
                     "Region Test" = percent(round(c(region_lda_test_error, region_qda_test_error, region_log_test_error), 4)),
                     "Area Training" = percent(round(c(area_lda_train_error, area_qda_train_error, area_log_train_error), 4)),
                     "Area Test" = percent(round(c(area_lda_test_error, area_qda_test_error, area_log_test_error), 4)))
rownames(df_temp) = c("Linear Discriminant Analysis", "Quadratic Discriminant Analysis", "Logistic Regression")

xtable_temp = xtable(df_temp)
digits(xtable_temp)[2:5] = 3
align(xtable_temp) = c('l', rep('c', 4))
print(xtable_temp, type='html', sanitize.colnames.function = function(x) gsub('\\.', ' ', x))
```

&nbsp;  

Overall, the task of classifying olive oils by area seemed to be harder than classifying by region,
as the training and test error rates for area classification models tended to be higher.   

This makes sense given that there are only 3 different regions of origin represented in the dataset,
but 9 different areas of origin. There are likely several geographic ‘areas’ contained within each
region, and so area classification is a more specific task.   

The region classification error rates can be thought of as a lower bound on the area classification rates. 
Assuming that multiple areas comprise each region, it would be impossible for area classification to be more 
accurate than region classification given the same data.

&nbsp;  

###Region Analysis  

The first part of this report will focus on the task of classifying the olive oils into one of 3
possible regions of origin.  

Before building the models outlined above, it’s useful to look at pairwise and parallel coordinates plots 
of the data to visualize the class structure.  

In each of the plots below, the colors correspond to the 3 different regions of origin found in the dataset:  

```{r, echo=FALSE, fig.width=12, fig.height=8}
ggpairs(train, 
        columns=3:n_cols, 
        diag='blank', 
        colour='Region', 
        axisLabels='internal',
        lower=list(continuous='points'), 
        upper=list(continuous='blank')) + theme_bw()
ggparcoord(train, scale='globalminmax', columns=3:n_cols, groupColumn=1) +
  xlab('acid') +
  theme_bw()
```

Because the typical values of different acids vary so much, it’s difficult to gain any insight into
the class structure in terms of individual acids by looking at the parallel coordinates plot
containing all 8 acids.   

It makes more sense to create several parallel coordinate plots with acids grouped by typical values.    

The following graphs exclude Oleic acid, whose class structure can already be seen fairly clearly in 
the parallel coordinates plot above:    

```{r, echo=FALSE, fig.width=12, fig.height=6}
plot1 = ggparcoord(train, scale='globalminmax', columns=c(3:7), groupColumn=1) +
  xlab('acid') +
  theme_bw()
plot2 = ggparcoord(train, scale='globalminmax', columns=c(4:5, 8:10), groupColumn=1) +
  xlab('acid') +
  theme_bw()
grid.arrange(plot1, plot2, nrow=1, ncol=2)
```


From these plots, it seems that Eicosenoic acid content clearly determines whether or not an
olive oil is from region 1, as olive oils that are not from region 1 contain almost no Eicosenoic
acid. In addition, it seems that olive oils from regions 2 and 3 have distinctly different Oleic
acid content. Using these two variables, Eicosenoic and Oleic acid content, we should be able
to build models capable of classifying the region a given olive oil is from with high accuracy.
The analysis that follows focuses on these two variables.  

Fitting a linear discriminant model to classify region of origin using the training data set and
the Eicosenoic and Oleic variables, we get the following prior probabilities and group means:  

```{r, echo=FALSE, results='asis'}
prior_probs = c(round(length(region_train_actual[region_train_actual==1])/n_train, 2),
                round(length(region_train_actual[region_train_actual==2])/n_train, 2),
                round(length(region_train_actual[region_train_actual==3])/n_train, 2))
oleic_mean = c(round(mean(train$Oleic[train$Region==1]), 2),
               round(mean(train$Oleic[train$Region==2]), 2),
               round(mean(train$Oleic[train$Region==3]), 2))
eicosenoic_mean = c(round(mean(train$Eicosenoic[train$Region==1]), 2),
                    round(mean(train$Eicosenoic[train$Region==2]), 2),
                    round(mean(train$Eicosenoic[train$Region==3]), 2))
df_temp = t(data.frame("Prior Probability" = prior_probs, 
                       "Oleic Mean" = oleic_mean, 
                       "Eicosenoic Mean" = eicosenoic_mean))
colnames(df_temp) = c("Region 1", "Region 2", "Region 3")
xtable_temp = xtable(df_temp)
align(xtable_temp) = c('l', rep('r', 3))
print(xtable_temp, type='html', sanitize.colnames.function = function(x) gsub('\\.', ' ', x),
                                sanitize.rownames.function = function(x) gsub('\\.', ' ', x))
```

&nbsp;  

In the training set, 57% of the olive oils were from region 1, while 16% were from region 2 and 
26% from region 3. As suspected, region 1 olive oils have a mean Eicosenoic content much greater 
than region 2 and region 3 olive oils. Region 3 olive oils have a mean Oleic acid content much greater 
than region 1 and region 2 olive oils.  

The classification rule seems to work well, as there are very few olive oils in the training or test
datasets incorrectly classified, as seen in the tables below:  

*LDA Training Classification Table:*  

```{r, echo=FALSE, results='asis'}
xtable_temp = xtable(table(region_train_lda_preds, region_train_actual))
colnames(xtable_temp) = c("Region 1 Actual", "Region 2 Actual", "Region 3 Actual")
rownames(xtable_temp) = c("Region 1 Predicted", "Region 2 Predicted", "Region 3 Predicted")
align(xtable_temp) = c('l', rep('c', 3))
print(xtable_temp, type='html', sanitize.colnames.function = function(x) gsub('\\.', ' ', x),
                                sanitize.rownames.function = function(x) gsub('\\.', ' ', x))
```

&nbsp;  

*LDA Test Classification Table:*  

```{r, echo=FALSE, results='asis'}
xtable_temp = xtable(table(region_test_lda_preds, region_test_actual))
colnames(xtable_temp) = c("Region 1 Actual", "Region 2 Actual", "Region 3 Actual")
rownames(xtable_temp) = c("Region 1 Predicted", "Region 2 Predicted", "Region 3 Predicted")
align(xtable_temp) = c('l', rep('c', 3))
print(xtable_temp, type='html', sanitize.colnames.function = function(x) gsub('\\.', ' ', x),
                                sanitize.rownames.function = function(x) gsub('\\.', ' ', x))
```

&nbsp;  

6 out of 505 training set olive oils and 1 out of 70 test set olive oil were incorrectly classified by
the LDA model, yielding a training error rate of 0.0119 and a test error rate of 0.0143.  

In the plot below, we can see the data projected onto the two linear discriminants:  

```{r, echo=FALSE, fig.width=12, fig.height=6}
LD_1_train = predict(lda_region_fit)$x[,1]
LD_2_train = predict(lda_region_fit)$x[,2]
plot1 = ggplot(data=train, aes(x=LD_1_train, y=LD_2_train, color=Region)) +
          geom_point(size=2) +
          labs(x='LD 1', y='LD 2', title='LDA Region - Training') +
          theme_bw()

LD_1_test = predict(lda_region_fit, newdata=test)$x[,1]
LD_2_test = predict(lda_region_fit, newdata=test)$x[,2]
plot2 = ggplot(data=test, aes(x=LD_1_test, y=LD_2_test, color=Region)) +
          geom_point(size=2) +
          labs(x='LD 1', y='LD 2', title='LDA Region - Test') +
          theme_bw()
grid.arrange(plot1, plot2, nrow=1, ncol=2)
```

Fitting a quadratic discriminant model to classify region of origin, again using the training
dataset and the Eicosenoic and Oleic variables, we get the following classification tables: 

*QDA Training Classification Table:*  

```{r, echo=FALSE, results='asis'}
xtable_temp = xtable(table(region_train_qda_preds, region_train_actual))
colnames(xtable_temp) = c("Region 1 Actual", "Region 2 Actual", "Region 3 Actual")
rownames(xtable_temp) = c("Region 1 Predicted", "Region 2 Predicted", "Region 3 Predicted")
align(xtable_temp) = c('l', rep('c', 3))
print(xtable_temp, type='html', sanitize.colnames.function = function(x) gsub('\\.', ' ', x),
                                sanitize.rownames.function = function(x) gsub('\\.', ' ', x))
```

&nbsp;  

*QDA Test Classification Table:*  

```{r, echo=FALSE, results='asis'}
xtable_temp = xtable(table(region_test_qda_preds, region_test_actual))
colnames(xtable_temp) = c("Region 1 Actual", "Region 2 Actual", "Region 3 Actual")
rownames(xtable_temp) = c("Region 1 Predicted", "Region 2 Predicted", "Region 3 Predicted")
align(xtable_temp) = c('l', rep('c', 3))
print(xtable_temp, type='html', sanitize.colnames.function = function(x) gsub('\\.', ' ', x),
                                sanitize.rownames.function = function(x) gsub('\\.', ' ', x))
```

&nbsp;  

The QDA model yielded a training error rate of 0.0099 and a test error rate of 0.0143. 5 out of
505 training set olive oils and 1 out of 70 test set olive oils were misclassified by the QDA
model, which is just a 1 training set olive oil improvement over the LDA model.  

Fitting a multivariate logistic regression model to classify region of origin, again using the
training dataset and the Eicosenoic and Oleic variables, we get the following classification
tables:  

*Logistic Regression Training Classification Table:*  

```{r, echo=FALSE, results='asis'}
xtable_temp = xtable(table(region_train_log_preds, region_train_actual))
colnames(xtable_temp) = c("Region 1 Actual", "Region 2 Actual", "Region 3 Actual")
rownames(xtable_temp) = c("Region 1 Predicted", "Region 2 Predicted", "Region 3 Predicted")
align(xtable_temp) = c('l', rep('c', 3))
print(xtable_temp, type='html', sanitize.colnames.function = function(x) gsub('\\.', ' ', x),
                                sanitize.rownames.function = function(x) gsub('\\.', ' ', x))
```

&nbsp;  

*Logistic Regression Test Classification Table:*  

```{r, echo=FALSE, results='asis'}
xtable_temp = xtable(table(region_test_log_preds, region_test_actual))
colnames(xtable_temp) = c("Region 1 Actual", "Region 2 Actual", "Region 3 Actual")
rownames(xtable_temp) = c("Region 1 Predicted", "Region 2 Predicted", "Region 3 Predicted")
align(xtable_temp) = c('l', rep('c', 3))
print(xtable_temp, type='html', sanitize.colnames.function = function(x) gsub('\\.', ' ', x),
                                sanitize.rownames.function = function(x) gsub('\\.', ' ', x))
```

&nbsp;  

The multivariate logistic regression model classified 3 out of 505 training olive oils and 1 out of
70 test olive oils incorrectly, yielding a training error rate of 0.0059 and a test error rate of
0.0143.  

&nbsp;  

###Area Analysis  

As we did earlier, before building any models we look at pairwise and parallel coordinates plots
of the data to visualize the class structure. In the plots below, the colors correspond to the nine
different areas of origin found in the dataset:  

```{r, echo=FALSE, fig.width=12, fig.height=8}
ggpairs(train,
        columns=3:n_cols,
        diag='blank',
        color='Area',
        axisLabels='internal',
        lower=list(continuous='points'),
        upper=list(continuous='blank')) + theme_bw()
ggparcoord(train,
           scale='globalminmax',
           columns=3:n_cols,
           groupColumn=2) + xlab('acid') + theme_bw()
```

```{r, echo=FALSE, fig.width=12, fig.height=6}
plot1 = ggparcoord(train,
                   scale='globalminmax',
                   columns=3:n_cols,
                   groupColumn=2) + xlab('acid') + theme_bw()
plot2 = ggparcoord(train,
                   scale='globalminmax',
                   columns=c(4:5, 8:10),
                   groupColumn=2) + xlab('acid') + theme_bw()
grid.arrange(plot1, plot2, nrow=1, ncol=2)
```

&nbsp;  

Fitting an LDA model using the training dataset and all acids to predict the Area of origin for
each olive oil, we get the following classification table (actual area in rows, predicted area in columns):

*LDA Training Classification Table:*  

```{r, echo=FALSE, results='asis'}
xtable_temp = xtable(table(area_train_lda_preds, area_train_actual))
colnames(xtable_temp) = c('Area 1', 'Area 2', 'Area 3', 'Area 4', 'Area 5', 'Area 6', 'Area 7', 'Area 8', 'Area 9')
rownames(xtable_temp) = c('Area 1', 'Area 2', 'Area 3', 'Area 4', 'Area 5', 'Area 6', 'Area 7', 'Area 8', 'Area 9')
align(xtable_temp) = c('l', rep('c', 9))
print(xtable_temp, type='html', sanitize.colnames.function = function(x) gsub('\\.', ' ', x),
                                sanitize.rownames.function = function(x) gsub('\\.', ' ', x))
```

&nbsp;  

*LDA Test Classification Table:*  

```{r, echo=FALSE, results='asis'}
xtable_temp = xtable(table(area_test_lda_preds, area_test_actual))
colnames(xtable_temp) = c('Area 1', 'Area 2', 'Area 3', 'Area 4', 'Area 5', 'Area 6', 'Area 7', 'Area 8', 'Area 9')
rownames(xtable_temp) = c('Area 1', 'Area 2', 'Area 3', 'Area 4', 'Area 5', 'Area 6', 'Area 7', 'Area 8', 'Area 9')
align(xtable_temp) = c('l', rep('c', 9))
print(xtable_temp, type='html', sanitize.colnames.function = function(x) gsub('\\.', ' ', x),
                                sanitize.rownames.function = function(x) gsub('\\.', ' ', x))
```

&nbsp;  

Out of the 505 training olive oils, 28 were classified into the incorrect area of origin by the
LDA model. This corresponds to a training error rate of 0.0554. Out of the 70 test set olive
oils, the LDA model incorrectly classified 7 olive oils, corresponding to a test error rate of 0.10.  

In the plot below, we can see the data projected onto the first two linear discriminants:  

```{r, echo=FALSE, fig.width=12, fig.height=6}
LD_1_train = predict(lda_area_fit)$x[,1]
LD_2_train = predict(lda_area_fit)$x[,2]
plot1 = ggplot(data=train, aes(x=LD_1_train, y=LD_2_train, color=Area)) +
          geom_point(size=2) +
          labs(x='LD 1', y='LD 2', title='LDA Area = Training') +
          theme_bw()
LD_1_test = predict(lda_area_fit, newdata=test)$x[,1]
LD_2_test = predict(lda_area_fit, newdata=test)$x[,2]
plot2 = ggplot(data=test, aes(x=LD_1_test, y=LD_2_test, color=Area)) +
          geom_point(size=2) +
          labs(x='LD 1', y='LD 2', title='LDA Area = Test') +
          theme_bw()
grid.arrange(plot1, plot2, nrow=1, ncol=2)
```

&nbsp;  

Fitting a QDA model using the training dataset and all eight acids to predict the area of origin
for each olive oil, we get the following classification tables:  

*QDA Training Classification Table:*  

```{r, echo=FALSE, results='asis'}
xtable_temp = xtable(table(area_train_qda_preds, area_train_actual))
colnames(xtable_temp) = c('Area 1', 'Area 2', 'Area 3', 'Area 4', 'Area 5', 'Area 6', 'Area 7', 'Area 8', 'Area 9')
rownames(xtable_temp) = c('Area 1', 'Area 2', 'Area 3', 'Area 4', 'Area 5', 'Area 6', 'Area 7', 'Area 8', 'Area 9')
align(xtable_temp) = c('l', rep('c', 9))
print(xtable_temp, type='html', sanitize.colnames.function = function(x) gsub('\\.', ' ', x),
                                sanitize.rownames.function = function(x) gsub('\\.', ' ', x))
```

&nbsp;  

*QDA Test Classification Table:*  

```{r, echo=FALSE, results='asis'}
xtable_temp = xtable(table(area_test_qda_preds, area_test_actual))
colnames(xtable_temp) = c('Area 1', 'Area 2', 'Area 3', 'Area 4', 'Area 5', 'Area 6', 'Area 7', 'Area 8', 'Area 9')
rownames(xtable_temp) = c('Area 1', 'Area 2', 'Area 3', 'Area 4', 'Area 5', 'Area 6', 'Area 7', 'Area 8', 'Area 9')
align(xtable_temp) = c('l', rep('c', 9))
print(xtable_temp, type='html', sanitize.colnames.function = function(x) gsub('\\.', ' ', x),
                                sanitize.rownames.function = function(x) gsub('\\.', ' ', x))
```

&nbsp;  

Out of the 505 training olive oils, only 5 were incorrectly classified by the QDA model, while
out of the 70 test olive oils, only 3 were incorrectly classified. This corresponds to a 0.0099
training error rate and a 0.0429 test error rate – markedly better than the LDA model.  

Fitting a multivariate logistic regression model to classify area of origin, again using the
training dataset and all eight acids to predict the area of origin for each olive oil, we get the
following classification tables:  

*Logistic Regression Training Classification Table:*  

```{r, echo=FALSE, results='asis'}
xtable_temp = xtable(table(area_train_log_preds, area_train_actual))
colnames(xtable_temp) = c('Area 1', 'Area 2', 'Area 3', 'Area 4', 'Area 5', 'Area 6', 'Area 7', 'Area 8', 'Area 9')
rownames(xtable_temp) = c('Area 1', 'Area 2', 'Area 3', 'Area 4', 'Area 5', 'Area 6', 'Area 7', 'Area 8', 'Area 9')
align(xtable_temp) = c('l', rep('c', 9))
print(xtable_temp, type='html', sanitize.colnames.function = function(x) gsub('\\.', ' ', x),
                                sanitize.rownames.function = function(x) gsub('\\.', ' ', x))
```

&nbsp;  

*Logistic Regression Test Classification Table:*  

```{r, echo=FALSE, results='asis'}
xtable_temp = xtable(table(area_test_log_preds, area_test_actual))
colnames(xtable_temp) = c('Area 1', 'Area 2', 'Area 3', 'Area 4', 'Area 5', 'Area 6', 'Area 7', 'Area 8', 'Area 9')
rownames(xtable_temp) = c('Area 1', 'Area 2', 'Area 3', 'Area 4', 'Area 5', 'Area 6', 'Area 7', 'Area 8', 'Area 9')
align(xtable_temp) = c('l', rep('c', 9))
print(xtable_temp, type='html', sanitize.colnames.function = function(x) gsub('\\.', ' ', x),
                                sanitize.rownames.function = function(x) gsub('\\.', ' ', x))
```

&nbsp;  

Out of the 505 training olive oils, the multivariate logistic regression model incorrectly
classified 82 olive oils and out of 70 test olive oils the model incorrectly classified 20 olive oils.
This corresponds to a training error rate of 0.1624 and a test error rate of 0.2857.  


