---
title: "Principal Component Analysis"
author: "Liam Abbott"
output: html_document
---

##Summary

The wine dataset contains the values of 13 different chemical characteristics for each of
180 different wines. After exploring the dataset using principal components analysis, it
seems that the first five principal components are the most important. Many of the
variables are almost entirely in the direction of the first principal component, while
several others are in the direction of principal component two.

##Analysis

Before beginning principal components analysis on the dataset, it’s useful to visualize
the relationships between different variables by looking at a few plots of the data. While
not an exhaustive collection of all possible plots between pairs of variables, the plots
shown below help visualize the relationships between some of the more strongly
correlated pairs of variables in the dataset:

```{r, echo=FALSE, warning=FALSE, message=FALSE}
list.of.packages = c("ggplot2",
                     "gridExtra",
                     "xtable",
                     "devtools",
                     "Rcpp")
new.packages = list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) {
  install.packages(new.packages)
}
for (i in 1:length(list.of.packages)) {
  require(list.of.packages[i], character.only=TRUE)
}
install_github("vqv/ggbiplot")
require(ggbiplot)
```

```{r, echo=FALSE}
wine = read.table('wine.txt', header=T, sep="")
plot1 = ggplot(wine, aes(x=Falavanoids, y=Total.phenols)) + 
          geom_point(col="blue") + 
          labs(x='Flavanoids', y='Total Phenols',title='Total Phenols vs. Flavanoids \n r = 0.8680') +
          theme_bw()
plot2 = ggplot(wine, aes(x=OD280.OD315, y=Total.phenols)) + 
          geom_point(col="blue") + 
          labs(x='OD280/OD315', y='Total Phenols',title='Total Phenols vs. OD280/OD315 \n r = 0.7056') +
          theme_bw()
plot3 = ggplot(wine, aes(x=Proline, y=Alcohol.content)) + 
          geom_point(col="blue") +
          labs(x='ProlineFGH', y='Alcohol', title='Alcohol vs. ProlineFGH \n r = 0.6449') +
          theme_bw()
plot4 = ggplot(wine, aes(x=OD280.OD315, y=Falavanoids)) +
          geom_point(col="blue") +
          labs(x='OD280/OD315',y='Flavanoids',title='Flavanoids vs. OD280/OD315 \n r = 0.7912') +
          theme_bw()
grid.arrange(plot1, plot2, ncol=2)
grid.arrange(plot3, plot4, ncol=2)
```

To get an initial understanding of the center and spread of each of the 13 variables in
the dataset, it’s useful to look briefly at the mean and standard deviation of each:    

```{r, echo=FALSE, results='asis'}
df_table = data.frame("Variable" = colnames(wine), "Mean" = apply(wine, 2, mean), "Standard Deviation" = apply(wine, 2, sd))
xtable_temp = xtable(df_table)
digits(xtable_temp)[c(1, 3)] = 2
align(xtable_temp) = rep("c", 4)
print(xtable_temp, type='html', include.rownames=FALSE,
      sanitize.colnames.function = function(x) gsub("\\.", " ", x))  
```
    
&nbsp;  

As the summary figures above show, the variables in the dataset seem to be measured
on different scales and likely with different units.   

For this reason, principal components analysis on correlations makes the most sense in 
this situation. It’s unlikely that there is any valuable information in the variance of 
the variables that is lost by applying PCA to the correlation matrix rather than the 
covariance matrix.

Performing PCA on correlations on this dataset produced 13 principal components, each
of which explained the following proportions of total variation in the data:  

```{r, echo=FALSE, results='asis'}
pca_out = princomp(wine, cor=TRUE)
eig_vals = pca_out$sdev^2
pve = eig_vals/sum(eig_vals)
df_table = data.frame("Principal Component" = seq(1:13), "Percentage of Variation Explained" = pve*100)
xtable_temp = xtable(df_table)
digits(xtable_temp)[2] = 2
align(xtable_temp) = rep("c", 3)
print(xtable_temp, type='html', include.rownames=FALSE,
      sanitize.colnames.function = function(x) gsub("\\.", " ", x))   
```

&nbsp;  

To visualize the information in the table above about the proportion of variation
explained by each principal component, we can look at a scree plot showing the variance
of each principal component on the y-axis:

```{r, echo=FALSE}
scree_plot = ggplot(data.frame(eig_vals = eig_vals, pc = seq(1:13)), aes(x=pc,y=eig_vals)) +
              geom_point(shape=16, size=3, col="blue") + 
              geom_line(col="darkgreen") +
              labs(x='Principal Component', y='Variance of Component', title='Scree Plot') +
              scale_x_continuous(breaks = seq(1, 13, 1)) +
              scale_y_continuous(breaks = seq(0, 5, 0.5)) +
              theme_bw()
scree_plot
```

We can look at the same information in a different way by plotting the proportion of
total variance in the dataset explained by each principal component and the cumulative
proportion of variance explained up to each principal component in the plots below:  

```{r, echo=FALSE}
df_temp = data.frame(pc=seq(1:13), pve=pve, cpve=cumsum(pve))
plot1 = ggplot(df_temp, aes(x=pc, y=pve)) +
          geom_point(shape=16, size=3, col="blue") + 
          geom_line(col="darkgreen") +
          ylim(0,1) +
          labs(x='Principal Component', y='Proportion of Variation Explained') +
          theme_bw()
plot2 = ggplot(df_temp, aes(x=pc, y=cpve)) + 
          geom_point(shape=16,size=3, col="blue") +
          geom_line(col="darkgreen") +
          ylim(0,1) +
          labs(x='Principal Component', y='Cumulative Proportion of Variation Explained') +
          theme_bw()
grid.arrange(plot1, plot2, ncol=2)
```

Looking at both the table and plots above, it seems that the first five principal
components are the most important in this dataset. Together, they combine to explain
about 80% of the variation in the data.  Going forward, we will focus the analysis on
these first five principal components.

After narrowing the focus to the first five principal components of the data, we can look
more closely at the variable loadings of each of these five principal components:

```{r, echo=FALSE, results='asis'}
loadings = data.frame(round(pca_out$loadings[,1:5],4))
colnames(loadings) = c("Prin Comp 1", "Prin Comp 2", 
                       "Prin Comp 3", "Prin Comp 4",
                       "Prin Comp 5")
xtable_loadings = xtable(loadings)
digits(xtable_loadings)[2:6] = 4
align(xtable_loadings) = c('l', rep("c", 5))
print(xtable_loadings, type='html')
```

&nbsp;  

From the table above, the first principal component draws fairly equally on many of the
variables, including malic acid, alkalinity of ash, total phenols, flavonoids, nonflavanoid
phenols, proanthocyanins, hue, OD280/OD315, and prolineFGH.  

Ash and color intensity are the only two variables that do contribute meaningfully to principal
component 1. On the other hand, these two variables are the primary contributors to
principal component 2, while several other variables are not much of a factor, including
alkalinity of ash, total phenols, flavonoids, nonflavanoid phenols, and proanthocyanins.  

Below, we can look at a plot of the data projected onto the first two principal
components:  

```{r, echo=FALSE}
qplot(x=pca_out$scores[,1], y=pca_out$scores[,2], data=wine, label=row.names(wine), geom='text') +
  theme_bw() +
  labs(x='Principal Component 1', y='Principal Component 2')
```


The data generally follows a parabolic shape when projected onto the plot with principal
component 1 on the horizontal axis and principal component 2 on the vertical axis.  

Wines 81 and 116 in the dataset seem like they may be outliers, as does wine 159. These
three points stray from the general pattern the rest of the data seems to be
following.  

One final plot to look at in this principal component analysis is a biplot showing the
original 13 variables projected onto a plot of the first two principal components:  

```{r, echo=FALSE}
ggbiplot(pca_out, labels=rownames(wine)) + theme_bw()
```

From looking at the biplot, the variables ‘total phenols’, ‘flavanoids’, ‘proanthocyanins’,
‘nonflavanoid phenols’, and ‘alcalinity of ash’ are almost entirely along the first
principal component. Meanwhile, ‘ash’, ‘alcohol’, and ‘color intensity’ are mostly along
the second principal component.
