---
title: "Principal Component Analysis"
author: "Liam Abbott"
date: "September 27, 2015"
output: html_document
---

##Summary

The wine dataset contains the values of 13 different chemical characteristics for each of
180 different wines. After exploring the dataset using principal components analysis, it
seems that the first five principal components are the most important. Many of the
variables are almost entirely in the direction of the first principal component, while
several others are in the direction of principal component two.

##Analysis

Before beginning principal components analysis on the dataset, it’s useful to visualize
the relationships between different variables by looking at a few plots of the data. While
not an exhaustive collection of all possible plots between pairs of variables, the plots
shown below help visualize the relationships between some of the more strongly
correlated pairs of variables in the dataset:

```{r, echo=FALSE, warning=FALSE, message=FALSE}
list.of.packages = c("ggplot2",
                     "gridExtra",
                     "xtable")
new.packages = list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) {
  install.packages(new.packages)
}
for (i in 1:length(list.of.packages)) {
  require(list.of.packages[i], character.only=TRUE)
}
```

```{r, echo=FALSE}
wine = read.table('wine.txt', header=T, sep="")
plot1 = ggplot(wine, aes(x=Falavanoids, y=Total.phenols)) + 
          geom_point(col="blue") + 
          labs(x='Flavanoids', y='Total Phenols',title='Total Phenols vs. Flavanoids \n r = 0.8680') +
          theme_bw()
plot2 = ggplot(wine, aes(x=OD280.OD315, y=Total.phenols)) + 
          geom_point(col="blue") + 
          labs(x='OD280/OD315', y='Total Phenols',title='Total Phenols vs. OD280/OD315 \n r = 0.7056') +
          theme_bw()
plot3 = ggplot(wine, aes(x=Proline, y=Alcohol.content)) + 
          geom_point(col="blue") +
          labs(x='ProlineFGH', y='Alcohol', title='Alcohol vs. ProlineFGH \n r = 0.6449') +
          theme_bw()
plot4 = ggplot(wine, aes(x=OD280.OD315, y=Falavanoids)) +
          geom_point(col="blue") +
          labs(x='OD280/OD315',y='Flavanoids',title='Flavanoids vs. OD280/OD315 \n r = 0.7912') +
          theme_bw()
grid.arrange(plot1, plot2, ncol=2)
grid.arrange(plot3, plot4, ncol=2)
```

To get an initial understanding of the center and spread of each of the 13 variables in
the dataset, it’s useful to look briefly at the mean and standard deviation of each:

```{r, echo=FALSE, results='asis'}
df_table = data.frame("Variable" = colnames(wine), "Mean" = apply(wine, 2, mean), "Standard Deviation" = apply(wine, 2, sd))
xtable_temp = xtable(df_table)
digits(xtable_temp)[c(1, 3)] = 2
print(xtable_temp, type='html', include.rownames=FALSE)
```

