---
title: 'Factor Analysis and Multidimensional Scaling'
author: "Liam Abbott"
output: html_document
---

###Summary  

The wine dataset contains 13 different chemical composition measurements for 180
wines. In this analysis, we use factor analysis and multidimensional scaling to reduce the
dimensionality of the dataset and assess the quality-of-fit of these lower-dimensional
representations.   

The multidimensional scaling approach results in a model containing
just two dimensions, while the factor analysis approach results in a model containing six
different factors. Compare these to the five components used in a previous principal
components analysis of this dataset.    

```{r, echo=FALSE, warning=FALSE, message=FALSE}
list.of.packages = c("ggplot2",
                     "gridExtra",
                     "xtable",
                     "cluster",
                     "stats")
new.packages = list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) {
  install.packages(new.packages)
}
for (i in 1:length(list.of.packages)) {
  require(list.of.packages[i], character.only=TRUE)
}
```

###Factor Analysis  

Before beginning factor analysis, look at the correlation matrix of the variables in the
wine dataset:  

```{r, echo=FALSE, results='asis'}
wine = read.table('wine.txt', header=TRUE, sep="")
wine_cov = round(var(wine), 2)
wine_cor = round(cor(wine), 2)
df_temp = data.frame(wine_cor)
df_temp[upper.tri(df_temp)] = NA
xtable_temp = xtable(df_temp)
digits(xtable_temp)[2:14] = 2
align(xtable_temp) = c('l', rep('c', 13))
print(xtable_temp, type='html', sanitize.colnames.function = function(x) seq(1:13))
```

&nbsp;  

For convenience, the column names in the correlation matrix above have been
compressed (1 = Alcohol, 2 = Malic acid, etc.). While some of the correlations are weak,
it seems that there are enough strong correlations between variables to make factor
analysis a reasonable undertaking.  

The first step is to decide how many factors should be included in the model. Assuming
normality, we can do this by looking at the likelihood ratio statistic for models fit with
different numbers of factors. In the previous principal components analysis of the wine
dataset, 5 principal components were included in the analysis based on the variance
explained by each component.  

Looking at the scree plot below, it seems reasonable to use 5 factors as a starting point
here:  

```{r, echo=FALSE}
eig_vals = eigen(wine_cor)$values[1:8]
scree_plot = ggplot(data.frame(eig_vals=eig_vals, factors=seq(1:8)), aes(x=factors, y=eig_vals)) +
              geom_point(shape=16, size=3, col="blue") +
              geom_line(col="darkgreen") + 
              labs(x='Factor', y='Eigenvalues', title='Scree Plot') +
              theme_bw() +
              scale_x_discrete(breaks=seq(1:8)) +
              scale_y_continuous(breaks=seq(0,5,0.5))
scree_plot
```

Fitting a factor analysis model with 5 factors to the wine dataset results in a model with
a likelihood ratio statistic of 39.75 on 23 degrees of freedom. The p-value of 0.0164
indicates that the model with only 5 factors is perhaps a bit too restrictive.  

A factor analysis model with 6 factors included results in a likelihood ratio statistic of
17.79 on 15 degrees of freedom. The p-value of 0.274 indicates that this model fits the
dataset fairly well. The variable loadings for each of the six factors are shown below:  

```{r, echo=FALSE, results='asis'}
wine_FA_5 = factanal(factors=5, x=wine)
wine_FA_6 = factanal(factors=6, x=wine)
df_temp = wine_FA_6$loading[1:13, 1:6]
xtable_temp = xtable(df_temp)
digits(xtable_temp)[2:7] = 2
align(xtable_temp) = c('l', rep('c', 6))
print(xtable_temp, type='html', sanitize.colnames.function = function(x) gsub('r', 'r\ ', x))
```

&nbsp;  

The loadings in the table above are from the factor analysis model using the varimax
rotation. As is clear from the loadings, for this dataset the varimax rotation does not
make an interpretation of the factors easy, as each factor draws on many of the
variables fairly significantly.   

To help interpret the loadings, we can fit another factor
analysis model with the promax rotation to get the following loadings:  

```{r, echo=FALSE, results='asis'}
wine_FA_6 = factanal(factors=6, x=wine, rotation='promax')
df_temp = wine_FA_6$loading[1:13, 1:6]
xtable_temp = xtable(df_temp)
digits(xtable_temp)[2:7] = 2
align(xtable_temp) = c('l', rep('c', 6))
print(xtable_temp, type='html', sanitize.colnames.function = function(x) gsub('r', 'r\ ', x))
```

&nbsp;  

Very roughly, it seems that each factor is drawing on these variables for the most part:  

```{r, echo=FALSE, results='asis'}
df_temp = data.frame(matrix(c(NA, 'x', 'x', NA, NA, NA,
                       NA, NA, NA, NA, 'x', NA,
                       NA, NA, NA, 'x', NA, NA,
                       NA, 'x', NA, 'x', NA, NA,
                       NA, NA, NA, NA, NA, 'x',
                       'x', NA, NA, NA, NA, NA,
                       'x', NA, NA, NA, NA, NA,
                       'x', NA, NA, 'x', NA, NA,
                       'x', NA, NA, NA, NA, NA,
                       NA, 'x', 'x', NA, NA, NA,
                       NA, NA, 'x', NA, 'x', NA,
                       'x', NA, 'x', NA, NA, NA,
                       NA, 'x', NA, NA, NA, NA),
                     nrow=13, ncol=6, byrow=TRUE))
colnames(df_temp) = c('Factor1', 'Factor2', 'Factor3', 'Factor4', 'Factor5', 'Factor6')
xtable_temp = xtable(df_temp)
align(xtable_temp) = c('l', rep('c', 6))
print(xtable_temp, type='html', sanitize.colnames.function = function(x) gsub('r', 'r\ ', x))
```

&nbsp;  

The 6 factors combine to explain about 77.9% of the total variance in the dataset, as
shown in the chart below:  

```{r, echo=FALSE, results='asis'}
df_temp = data.frame(matrix(c(0.26, 0.14, 0.10, 0.10, 0.09, 0.08,
                              0.26, 0.40, 0.50, 0.60, 0.69, 0.77),
                            nrow=2, ncol=6))
colnames(df_temp) = c('Factor1', 'Factor2', 'Factor3', 'Factor4', 'Factor5', 'Factor6')
rownames(df_temp) = c('Proportion of Variation Explained', 'Cumulative Proportion of Variation Explained')
xtable_temp = xtable(df_temp)
align(xtable_temp) = c('l', rep('c', 6))
print(xtable_temp, type='html', sanitize.colnames.function = function(x) gsub('r', 'r\ ', x))
```

&nbsp;  

We can visualize the information in the chart by graphing the variance explained versus
the factors:  

```{r, echo=FALSE}
pve = c(0.26, 0.14, 0.10, 0.10, 0.09, 0.08)
df_temp = data.frame(factors=seq(1:6), pve=pve, cpve=cumsum(pve))
plot1 = ggplot(df_temp, aes(x=factors, y=pve)) +
          ylim(0,1) +
          labs(x='Factors', y='Proportion of Variation Explained') +
          geom_point(shape=16, size=3, col="blue") +
          geom_line(col="darkgreen") +
          theme_bw() +
          scale_x_discrete(breaks=seq(1:6))
plot2 = ggplot(df_temp, aes(x=factors, y=cpve)) +
          ylim(0,1) +
          labs(x='Factors', y='Cumulative Proportion of Variation Explained') +
          geom_point(shape=16, size=3, col="blue") +
          geom_line(col="darkgreen") +
          theme_bw() +
          scale_x_discrete(breaks=seq(1:6))
grid.arrange(plot1, plot2, ncol=2)
```


###Multidimensional Scaling  

In addition to factor analysis, we can also apply a multidimensional scaling approach to
the wine dataset to see if the dissimilarities between the 13 variables in the dataset can
be represented in a lower dimensional space. Using Kruskal’s scale to assess the quality
of fit, a multidimensional scaling approach for one dimension up to six dimensions
produces the following graph of fit quality versus number of dimensions:

```{r, echo=FALSE}
wine_dist = daisy(wine, stand=TRUE)

wine_mds_1 = cmdscale(wine_dist, k=1)
fitted_dist_1 = daisy(wine_mds_1)
fit_qual_1 = sum((wine_dist - fitted_dist_1)^2)/sum(fitted_dist_1^2)

wine_mds_2 = cmdscale(wine_dist, k=2)
fitted_dist_2 = daisy(wine_mds_2)
fit_qual_2 = sum((wine_dist - fitted_dist_2)^2)/sum(fitted_dist_2^2)

wine_mds_3 = cmdscale(wine_dist, k=3)
fitted_dist_3 = daisy(wine_mds_3)
fit_qual_3 = sum((wine_dist - fitted_dist_3)^2)/sum(fitted_dist_3^2)

wine_mds_4 = cmdscale(wine_dist, k=4)
fitted_dist_4 = daisy(wine_mds_4)
fit_qual_4 = sum((wine_dist - fitted_dist_4)^2)/sum(fitted_dist_4^2)

wine_mds_5 = cmdscale(wine_dist, k=5)
fitted_dist_5 = daisy(wine_mds_5)
fit_qual_5 = sum((wine_dist - fitted_dist_5)^2)/sum(fitted_dist_5^2)

wine_mds_6 = cmdscale(wine_dist, k=6)
fitted_dist_6 = daisy(wine_mds_6)
fit_qual_6 = sum((wine_dist - fitted_dist_6)^2)/sum(fitted_dist_6^2)

fit = c(fit_qual_1, fit_qual_2, fit_qual_3, fit_qual_4, fit_qual_5, fit_qual_6)

qplot(x=seq(1:6), y=fit) + 
  labs(x='Dimensions', y='Fit Quality') +
  geom_point(shape=16, size=3, col="blue") +
  geom_line(col="darkgreen") +
  ylim(0, 1) +
  theme_bw()
```

While the quality of fit measure for 6 dimensions is an excellent 0.013 (lower the better in this case), 
it appears that just 2 dimensions is also adequate and has the advantage of being more easily visualized, 
as seen in the 2-dimensional representation of the wine dataset below:  

```{r, echo=FALSE}
qplot(x=wine_mds_2[,1], y=wine_mds_2[,2]) +
  geom_point(col="blue") +
  labs(x='First Dimension', y='Second Dimension') +
  theme_bw()
```
